Here we are going to analysis Loan prediction data set from analytics vidya.
This dataset has two part one is training data set(train.csv file) and another one is test data set(test.csv)
Let's upload them in work space.

```{r}
getwd()
setwd("C:/Users/Ashish Mishra/Desktop/DS/DataScienceProjects/Loan Prediction")
train <- read.csv("train.csv",stringsAsFactors = F)
#stringsAsFacors to convert factors into strings/char
test <- read.csv("test.csv",stringsAsFactors = F)
#We need to load some R packages for our analysis,let's  do that first
library(ggplot2)  # data exploration
library(ggthemes)  # data exploration
library(caret)    # predictive machine learning algorithem
library(dplyr)
library(randomForest)
full <- bind_rows(train,test)
```

First we need to analysis our data,let's see what  are the variables  we have in our training and test dataset

```{r}
str(full)
```
Our training data set has 13 variables,which includes variables credit history, applicant Income,education,employment etc., these are our predictors and Laon Status is target variable.

```{r}
str(test)
```
Test data set does not have Laon Status variable ,that we need to predict.

```{r}
summary(full)
sum(is.na(train))
sum(is.na(test))

```
We have NA  in train and test data. 
We can apply preprocessing  by using preprocess function in caret package,lets do that!!
```{r}
preObj<- preProcess(full[,-13], method=c("medianImpute","center", "scale"))

full_processed <- predict(preObj,full)
sum(is.na(full_processed))


```
Now we have only Loan_Status as missing values,which we need to predict.Thats why I have excluded Loan_Status variable in precprocessing.
here we  have replaced missing values by using median.


```{r}
set.seed(1)


train <- full_processed[1:614,]
test <- full_processed[615:981,]

summary(test)
```

```{r}
fitControl <- trainControl(method = "cv",number = 5,savePredictions = 'final',classProbs = T)
#Our predictors and outcome
predictors<-c("Credit_History", "LoanAmount", "Loan_Amount_Term", "ApplicantIncome","CoapplicantIncome")

outcome <- 'Loan_Status'
```

```{r}
#logistic model 
modFit <- train(train[,predictors],train[,outcome],model ='glm',trControl=fitControl,tuneLength = 3)

#prediction
test$Loan_Status<-predict(object = modFit,test[,predictors])

solution <- data.frame(Loan_ID = test$Loan_ID, Loan_Status = test$Loan_Status)

# Write the solution to file
write.csv(solution, file = 'Loan_pred_solution.csv', row.names = F)

```

On test data we got 0.77778 score on analytics vidya forum.

